Forecasting the power of Higher Order Weak Lensing Statistics with automatically differentiable simulations

### 2023-03-16 paper/paper.tex

% Add your own macros here:
\pdfoutput=1 %for arXiv submission
%\usepackage{amsmath,amssymb,amstext}
\usepackage{amsmath,amstext}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage{ae,aecompl}
\usepackage[utf8]{inputenc}
\usepackage[figure,figure*]{hypcap}

\usepackage{url}
\urlstyle{same}

\usepackage{lineno}
\linenumbers
%\modulolinenumbers[2]

@@

\newcommand{\EiffL}[1]{{\color{cyan}FL: #1}}
\newcommand{\denise}[1]{{\color{red}DL: #1}}

% header settings
\shorttitle{Differentiable Lensing Simulations}
\shortauthors{Lanzieri et al.\ (LSST~DESC)}

@@

\begin{abstract}
In this project we investigate the relative constraining power of various map-based higher order weak lensing statistics in an LSST setting. Such comparisons are typically very costly as they require a large number of simulations, and become intractable for more than 3 or 4 cosmological parameters. Instead we propose a new methodology based on the TensorFlow framework for automatic differentiation, and more specifically the public FlowPM N-body simulation code. By implementing ray-tracing in this framework, we will be able to simulate lensing lightcones, and compute Higher Order lensing statistics on the resulting maps. These statistics can then be differentiated with respect to cosmology, or any systematics included in the simulations, thus allowing, for exact Fisher forecasts.
\end{abstract}

@@

\section{Introduction}

\section{Methods}

%We compare the sensitivity of three different map-based statistics: angular power spectra, the peak counts and the $l_{1\text{norm}}$.
 %the calculation of the summary statistics, and the Fisher Forecast process.

\subsection{{Weak lensing}}

Weak gravitational lensing is a promising observational technique to infer the projected matter density distribution between an observer and a source by measuring galaxy shape correlations. 
The effect of gravitational lensing 
can be quantified in term of the separation vector \textbf{x} between a ray and a fiducial ray separated by the angle $\boldsymbol{\theta}$ at the observer:
\begin{align}\label{sep}
   & \textbf{x}(\boldsymbol{\theta},\chi) =
    f_k(\chi)\boldsymbol{\theta} + \\
   &- \frac{2}{c^2}
    \int_0^{\chi}
    \text{d}\chi'
    f_k(\chi-\chi')
    [ \boldsymbol{\nabla_{\bot}}\Phi( \textbf{x}(\boldsymbol{\theta},\chi'),\chi')-
    \boldsymbol{\nabla_{\bot}}\Phi^{(0)}(\chi')
    ],
\end{align}
where we denote as $\Phi$ and $\Phi^{0}$ the potential along the two light rays, and as $f_k(\chi)$ and $\chi$ the angular and radial comoving distance.

The Jacobian matrix, describing the linearized lens mapping, is defined as:
\begin{equation}
\mathcal{A}(\boldsymbol{\theta},\chi)=
\frac{1}{f_{k}(\chi)}
\frac{\partial \textbf{x}}{\partial \boldsymbol{\theta}}.
\end{equation}
We know that the knowledge of the gravitational potential and the use of accurate solvers for light ray trajectories is computationally expensive. Nevertheless, the integral equation \ref{sep} can be approximated by considering the series expansion in power of $\Phi$ and truncating the series at the first term.
With this assumptions, and considering that $\nabla_{\bot}\boldsymbol{\Phi}^0$ does not depend from $\boldsymbol{\theta}$, the Jacobian matrix can be written as:
\begin{equation}
    \mathcal{A}_{ij}(\boldsymbol{\theta},\chi)
     =\delta_{ij}-\frac{2}{c^2}
    \int_0^{\chi} d\chi'
     \frac{f_k(\chi-\chi')f_k(\chi')}{f_k(\chi)}
     \Phi_{ij}(f_k(\chi')\boldsymbol{\theta},\chi').
\end{equation}
Notice that this approximation, also known as Born approximation, is only valid in the limit of weak-field metric (small $\Phi$).

If we define the 2D potential, the \textit{lensing potential} as:
\begin{equation}
    \psi(\boldsymbol{\theta},\chi) \equiv
 -\frac{2}{c^2}
= \int_0^{\chi} d\chi'
   \frac{f_k(\chi-\chi')f_k(\chi')}{f_k(\chi)f_k(\chi')}
  \Phi(f_k(\chi')\boldsymbol{\theta},\chi').
\end{equation}
 the Jacobi matrix can be written as:
\begin{equation}
    \mathcal{A}_{ij}=\delta_{ij}-\partial_i \partial_j\psi. 
\end{equation}

From the parametrization of the symmetrical matrix $\mathcal{A}$, we can define the spin-two shear $\gamma=(\gamma_1,\gamma_2)$ and the scalar convergence field, $\kappa$. 
Hence, the convergence and the shear are defined as second derivative of the potential:
\begin{equation}\label{kshear}
    \kappa=\frac{1}{2}(\partial_1\partial_1+\partial_2\partial_2)\psi; \ \gamma_1=\frac{1}{2}(\partial_1\partial_1-\partial_2\partial_2)\psi; \ 
    \gamma_2=\partial_1\partial_2\psi;
\end{equation}

The two fields $\gamma$ and $\kappa$ describe the distortion in the shape of the image, and the change in the angular size, respectively.
By combining the 2D Poisson equation with the Eq.\ref{kshear} the convergence $\kappa$ becomes:
\begin{equation}
    \kappa_{born}(\boldsymbol{\theta})= \frac{3H_0^2 \Omega_m}{2c^2}
    \int_0^{\chi_s} 
    \frac{d\chi}{a(\chi)}
    g(\chi)
    \delta(f_k(\chi)\boldsymbol{\theta},\chi),
\end{equation}
where we define the \textit{lensing efficiency}
\begin{equation}
   g(\chi) \equiv
   \text{d}\chi'
   n(\chi')
    \frac{f_k(\chi'-\chi)}{f_k(\chi')}.
\end{equation}

@@

\subsection{Simulation}
It's well known that an common feature of statics beyond second-order is the absence of analytical models with which to predict the observed signals. Hence, we need to adopt a forward modelling approach and create a suite of numerical simulations. 
In the following section we introduce the mass map simulation procedure, including a description of the simulator and of the construction of the weak lensing lightcone.

\subsubsection{FLOWPM}
Our weak lensing simulation tool is based on FlowPM (/cite Modi, Lanusse et al. ), a fast particle-mesh solver that implements the time integration for gravitational evolution starting from the Gaussian initial conditions of the Universe to the observed large scale structures through a Kick-Drift scheme.

After instantiating the $\Lambda$CDM cosmology with our desired parameters, (the initial power spectrum and the evolution of the particles depend from $\sigma_8$ and $\Omega_{c}$) we created the initial condition for the N-body simulation using the normalised linear matter power spectrum as implemented by (/cite Eisenstein and Hu of '98 ) and generated the linear field at requested scale. After that, we implemented the N-body. As said before, The N-body simulation does the time integration of a given initial linear field  across a given array of scale factors implementing a series of Kick-Drift-Kick operations. 
During the \textit{Kick} operation we estimate the gravitational force experienced by every particles implementing Fast Fourier Transform and update their momentum. In the next state, \textit{Drift}, the particles are displaced and their positions is updated with the velocity estimated in the previous step. This operation is continued until the observed large scale structures at the final time-step.
To perform the lightcone computation and extract the lens planes,  we export several intermediate state from the N-body simulation. From a given intermediate state we extract a slice and project it as a density plane by painting the particles on a 2D mesh. 
After creating a grid of coordinates, the lens planes are interpolated onto sky coordinates. 
 At this point, we have constructed a ray tracer object, to compute the convergence, we directly integrate the lensing density along the unperturbed line of sight.

\subsubsection{Additional validation for our ray-tracing implementation } 
As an additional test to validate our ray-tracing, instead of running the N-body simulation to generate the snapshots of particles for different scale factors, we create these snapshots by using a linear matter power spectrum at a given fixed redshift $z=0.52$ . \\
Specifically, we estimate the initial LPT displacement given an input linear (real) field.

As we said, the aim of this test is to prove the validity of our ray-tracing. In principle, without the N-body simulations and his possible approximations and systematics, any disagreement between the angular power spectrum computed from the final convergence map
and the angular power spectrum we use as reference can be attributable to the ray-tracing.

Fig. shows a comparison between the theoretical linear matter power spectrum computed using the Jax-cosmo library at the redshift $0.52$ and the matter power spectrum computed from the resulting snapshots, after applying the painting operations. As we can see the matter power spectrum start to lack in power for $k$(maggiore o circa uguale)1. This is probably due to the paint operation and is something to take into account when we will study the final power spectrum.

From the full projected boxes of the simulations, we exported the lens planes, then interpolated on the light-cones.

The final angular power spectrum is obtained averaging over 1000 samples.

\subsubsection{Potential Gradient Descendent (PGD)}    
Fast N-body simulations can be used as viable alternative to full N-body to model the galaxy statistics and create fast and low computational costs realizations of the large scale structures. Nevertheless, this kind of simulations lack resolution on small scale and can't give accurate halo matter profiles or matter power spectrum.

To recover this missing accuracy, we employs the PGD scheme (\cite{Dai_2018}). 
The main goal is to mimic the physics lost in the low resolution simulations by computing the short range force operation as an additional correction displacements in a post-processing step. \\
The additional displacement is computed by pointing the particles towards a filtered potential minimum of the particle mesh solver.

%\begin{equation}\label{displacement}
 %   \textbf{S}=4\pi G \bar{\rho}\alpha_{PM}/H_0^2 \nabla(
  % \hat{\textbf{O}}_l(k)
   %  \hat{\textbf{O}}_s(k)
    %\nabla^{-2}\delta),
%\end{equation}
%where $\hat{\textbf{O}}_l(k)= \exp{(\frac{k^2}{k^4_l})}$ and $\hat{\textbf{O}}_s(k)= \exp{(-\frac{k^4}{k^4_s}})$ are the low and high pass filter introduced to remove the long range force and to reduce the numerical effect induced by the mesh resolution. \\
PGD introduces control parameters that can be tuned by comparing the matter power spectrum to a reference power spectrum.

We employ an online-PGD technique worth mentioning here: the correction involved in computing additional displacements is computed after every drift step in the N-body simulation.

We find that a PGD correction computed directly at the end of the simulation fails, introducing some significant sharp clusters or changing the shape of some of the structures. Effectively this is equivalent to requesting a too large correction at once, which our method avoids.

In practice we set it up so that it takes the power spectra at all intermediate redshift and fits the PGD parameters as to recover the \textit{Halofit} non linear power spectrum.

Figure \ref{fig:pkhalofit_comp} shows the matter power spectrum at $z=1.69$ of our simulations before and after applying the PGD correction compared to analytic Halofit predictions (\denise{cita}(Smith et al. 2003; Takahashi et al. 2012)).
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/pkhalofit_comp.png}
    \caption{
    Upper panel: Comparison of the matter power spectrum of FlowPM before and after the PGD correction with reference matter power spectrum \textit{Halofit} at redshift $z = 1.69$. Lower panel: Fractional differences between FlowPM and Halofit.
    }
    \label{fig:pkhalofit_comp}
\end{figure}

In Figure \ref{fig:comp_kmap} we show an example of our convergence map at $ z= 1.034$, as well as the same map when the PGD correction is not applied. 
We can see that the structures are not modified but simply sharpened.
\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{paper/figures/kmap_compgd.png}
    \caption{ \textbf{Left panel}: An example of our convergence map at source redshift $z= 1.034$.
    \textbf{Right panel}: Same convergence map when the PGD correction is not applied.}
    \label{fig:comp_kmap}
\end{figure*}

In Figure \ref{fig:ps_comp} in the upper panel, we present the DLL angular power spectrum complemented by the PGD scheme to a conventional DLL simulation with the same resolution. Both the outputs are compared to the analytic Halofit prediction for the angular power spectrum. In the lower panel the fractional differences between the our maps and the analytic Halofit prediction.  Both the power spectra and ratios are averaged over $N = 100$ realisations.
We can conclude that PGD scheme allows to achieve higher accuracy at the same computational cost.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{paper/figures/clshalofit_comp.png}
    \caption{
    \textbf{Upper panel}: Angular power spectra for 3 different source redshifts from
our DLL maps. \textbf{Lower panel}: Fractional differences between the
DLL and Halofit power spectra for 3 source redshifts. The power spectra and ratios are means over 100 map realisations. The two vertical lines define the range of $\ell$ between $[300,3000]$.}
    \label{fig:ps_comp}
\end{figure}

\subsection{Summary statistics}
In this work, in order to extract the cosmological information from the simulated $\kappa-$maps, we use tree summary statistic for weak lensing observable: the angular power spectrum, the `starlet peak counts' and the `starlet $l_{1\text{norm}}$'.

We reproduce the analysis choices of the LSST Y1 data. The DESC science requirements document can be found here \cite{mandelbaum2018lsst}.
In particular we use a $n_{\text{eff}}=10$ galaxies per $\text{arcmin}^2$ for lensing sources and a $\sigma_e=0.26$ per component.

\subsubsection{Lensing Angular Power Spectrum}
Second order statistics, both in the form of shear 2-point correlation function $\xi_{\pm}(\theta)$, or its counterpart in Fourier space, the angular power spectrum $C_{\ell}$, have been widely used to extract the cosmological information from weak lensing mass maps. \
In the Limber approximation, the angular power spectrum of the convergence field can be computed as:
\begin{equation}
    C_{\kappa}(\ell)=\frac{9\Omega_m^2H_0^4}{4c^4}
    \int_0^{\chi_{lim}} \text{d}\chi 
    \frac{g^2(\chi)}{a^2(\chi)}P_{\delta}
    	\left( k=\frac{\ell}{f_K(\chi)},\chi \right),
\end{equation}
where $P_{\delta}$ defines the matter power spectrum of the density contrast.

In this paper, the angular power spectrum is computed from the noisy smoothed convergence maps: to each mock $\kappa$map we add Gaussian noise and than convolved it with a Gaussian smoothing kernel.

>>>>

% Add your own macros here:
\pdfoutput=1 %for arXiv submission
%\usepackage{amsmath,amssymb,amstext}
\usepackage{amsmath,amstext}
\usepackage[T1]{fontenc}
\usepackage{apjfonts}
\usepackage{ae,aecompl}
\usepackage[utf8]{inputenc}
\usepackage[figure,figure*]{hypcap}
\usepackage{hyperref}
\usepackage{url}
\urlstyle{same}
\usepackage{lineno}
\linenumbers
%\modulolinenumbers[2]

@@

\newcommand{\jl}[1]{{\color{orange}JL: #1}}
\newcommand{\EiffL}[1]{{\color{cyan}FL: #1}}
\newcommand{\denise}[1]{{\color{red}DL: #1}}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{fontawesome}
\newcommand{\github}{\href{https://github.com/DifferentiableUniverseInitiative/flowpm/tree/master}{\faGithub}}
\newcommand{\nblink}[1]{\href{https://github.com/DifferentiableUniverseInitiative/flowpm/tree/nbody_ode/notebooks/#1.ipynb}{\faFileCodeO}}
% header settings
\shorttitle{Differentiable Lensing Simulations}
\shortauthors{Lanzieri et al.\ (LSST~DESC)}

@@

\begin{abstract}
We present the Differentiable Lensing Lightcone (DLL), a fully differentiable physical model designed for being used as a forward model in Bayesian inference algorithms requiring access to derivatives of lensing observables with respect to cosmological parameters. 
We extend the public FlowPM N-body code, a particle-mesh N-body solver, simulating lensing lightcones \denise{and implementing the Born approximation} in the Tensorflow framework. Furthermore, DLL is aimed at achieving high accuracy with low computational costs. As such, it integrates a novel Hybrid Physical-Neural parameterisation able to compensate for the small-scale approximations resulting from particle-mesh schemes for cosmological N-body simulations.
We validate our simulations in an LSST setting against high-resolution $\kappa$TNG simulations by comparing both the lensing angular power spectrum and multiscale peak counts. We demonstrate an ability to recover lensing $C_\ell$ up to a 10\% accuracy at $\ell=1000$ for sources at redshift 1, with as few as $\sim 0.6$ particles per Mpc/h. 
As a first use case, we use this tool to investigate the relative constraining power of the angular power spectrum and peak counts statistic in an LSST setting.
Such comparisons are typically very costly as they require a large number of simulations, and do not scale well with the increasing number of cosmological parameters.  
As opposed to forecasts based on finite differences, these statistics can be analytically differentiated with respect to cosmology, or any systematics included in the simulations at the same computational cost of the forward simulation. We find that the peak counts outperform the power spectrum on the cold dark matter parameter $\Omega_c$, on the amplitude of density fluctuations $\sigma_8$, and on the
amplitude of the intrinsic alignment signal $A_{IA}$.
\end{abstract}

@@

\section{Introduction}
Weak gravitational lensing by Large Scale Structures (LSS) is one of the key probes to test cosmological models and gain insight into constituents of the Universe. The upcoming stage-IV surveys, such as the Legacy Survey of Space and Time (LSST) of the Vera C. Rubin Observatory \citep{ivezic2019lsst}, the Nancy Grace Roman Space Telescope \citep{spergel2015wide}, and  the Euclid Mission \citep{laureijs2011euclid}, will provide measurements of billions of galaxy shapes with unprecedented accuracy, which in turn will lead to tight constraints on dark energy models \citep[e.g.][]{mandelbaum2018lsst}.

With the increased statistical power of these surveys comes the question of their optimal analysis. Traditional cosmological analysis rely on measurements of the two-point statistics, either the shear two-point correlation functions or its Fourier transform, the lensing power spectrum. However, the two-point statistics are only optimal for Gaussian fields, and do not fully capture the non-Gaussian information imprinted in the lensing signal at the scales that future surveys will be able to access (e.g. information encoded in the peaks and in the filamentary features of the matter distribution).

 This has led to the introduction of a number of  higher-order statistics to access the non-Gaussian information from weak lensing data: 
the weak lensing one point PDF \citep{liu2019constraining, uhlemann2020fisher, boyle2021nuw}, lensing peak counts \citep{liu2015cosmology, liu2015cosmological, lin2015new, kacprzak2016cosmology, peel2017cosmological, shan2018kids, martinet2018kids, ajani2020constraining, harnois2021cosmic, zurcher2022dark}, Minkowski functionals \citep{kratochvil2012probing, petri2013cosmology}, moments of mass maps \citep{gatti2021dark}, wavelet and scattering transforms \citep{ajani2021starlet, cheng2021weak},, and 3 point statistics \citep{takada2004cosmological, semboloni2011weak, rizzato2019tomographic, halder2021integrated}.

Recently, machine learning-based methods that broadly fall in the category of Simulation-Based Inference (SBI) \citep{fluri2019cosmological, kacprzak2022deeplss, fluri2021cosmological, jeffrey2021likelihood, fluri2022full}, and Bayesian forward-modeling frameworks \citep{porqueres2021bayesian, sarma2022map} have also been introduced to attempt to fully account for the non-Gaussian content in the weak lensing signal. Unlike the methods described above, these approaches are designed to access the full field-level information content. Even though these techniques are asymptotically theoretically optimal in terms of information recovery, they still suffer from significant limitations.

SBI methods are characterized by the absence of an analytical model to describe the observed signal and instead rely on learning a likelihood from simulations. Modern approaches employ deep learning-based density estimation methods to model the likelihood without the need to make any Gaussianity assumptions, and as such have 
drawn the attention of the community \citep{alsing2018massive, jeffrey2021likelihood}.

A key element common to most of these methodologies is their ability to benefit from gradient information. 
For example, \cite{porqueres2021bayesian} present a Bayesian hierarchical approach to infer the cosmic matter density field simultaneously with cosmological parameters using the Hamiltonian Monte Carlo (HMC) algorithm to explore the full high-dimensional parameter space. The HMC algorithm exploits the information encoded in the gradients of the joint likelihood function to inform each update step and hence requires the derivatives of the forward model.

 As a different class of examples, \cite{makinen2021lossless} demonstrate that the full information content of a cosmological field can be represented by optimal summaries, allowing for likelihood-free and near-exact posteriors for cosmological parameters. In particular, they use neural networks trained on simulations to maximise the Fisher information, which requires having access to derivatives of the simulation model to compute a Fisher matrix.

One possible way of evaluating these gradients with respect to cosmological parameters is by using numerical differentiation, e.g. computing a finite difference of a given statistic by varying the simulation fiducial values by a small amount.
However, numerical differentiation is expensive in terms of computational resources and simulation time, and also requires hyperparameter tuning for the step size used in the finite difference scheme. An alternative option consists in computing the gradient analytically. Both these solutions are faced with limitations: the first approach gets computationally intractable in high dimensions, while the analytic gradients are sometimes impossible to estimate.
All of these new techniques for cosmological analyses make the development of fast and differentiable simulations necessary.

\cite{bohm2021madlens} developed MADLens, a CPU-based python package for producing non-Gaussian lensing convergence maps. These maps are differentiable with respect to the initial conditions of the underlying numerical simulations and to the cosmological parameters $\Omega_m$ and $\sigma_8$. In this paper,  we aim to efficiently compute gradients that benefit the development of new inference algorithms for weak lensing surveys. To do this we extend the framework of the FlowPM package \cite{modi2021flowpm} by implementing the Born approximation and simulating lensing lightcones in the Tensorflow framework. TensorFlow is a tensor library that includes the ability to perform automatic differentiation. Automatic differentiation enables us to compute gradients exactly as opposed to finite differences, which only provide approximate gradients. Specifically, TensorFlow implements the backpropagation algorithm to compute gradients, i.e. first it creates a graph (e.g. data structures representing units of computation), then it works backward through the graph by applying the chain rule at each node. \denise{Unlike MADlens, our tool is GPU-based and provides derivatives with respect to all the cosmological parameters. There are also differences in the implementation of various functions to improve accuracy and speed.}

We validate our simulations against the cosmological N-body simulations $\kappa$TNG \citep{osato2021kappatng} by comparing both the lensing angular power spectrum and multiscale peak counts.

In particular, as a first application, we show how the differentiability of numerical simulations can be exploited to evaluate the Fisher Matrix. Then, we compare the constraining power of two map-based weak lensing statistics: the lensing power spectrum and peak counts and investigate the degeneracy in high dimensional cosmological and nuisance parameter space through Fisher forecasts.

This paper is structured as follows: in \autoref{weak_lesning_modeling} we briefly review the weak lensing modeling including the theoretical framework and the summary statistics used in this work. In \autoref{Fast_and_Differentiable_Lensing_Simulations} we introduce the numerical simulations illustrating the numerical methods used to generate mock WL maps. In \autoref{Validating_simulations_for_LSST_Y} we validate the simulations by comparing the statistics from our simulations and $\kappa$TNG-Dark ones.
The Fisher forecast formalism and the survey and noise setting are shown in \autoref{Application_Fisher_forecast_for_HOS}. We finally discuss our results and present our conclusions afterward, in \autoref{Discussion} and \autoref{Conclusions}.

\section{Weak Lensing Modeling}\label{weak_lesning_modeling}
\subsection{{Cosmic shear}}
Weak gravitational lensing is a powerful probe to infer the projected matter density distribution between an observer and a source.
The effect of gravitational lensing 
can be quantified in term of the separation vector \textbf{x} between two light rays separated by an angle $\boldsymbol{\theta}$:
\begin{align}\label{sep}
   & \textbf{x}(\boldsymbol{\theta},\chi) =
    f_k(\chi)\boldsymbol{\theta} + \\
   - \frac{2}{c^2}
   \int_0^{\chi}
    &  \text{d}\chi'
    f_k(\chi-\chi')
    [ \boldsymbol{\nabla_{\bot}}\Phi( \textbf{x}(\boldsymbol{\theta},\chi'),\chi')-
    \boldsymbol{\nabla_{\bot}}\Phi^{(0)}(\chi')
    ],
\end{align}
with $\Phi$ and $\Phi^{0}$ the gravitational potential along the two light rays, $f_k(\chi)$ and $\chi$ the angular and radial comoving distance.

Formally, the effect of the linearized lens mapping is described by the Jacobian matrix:
\begin{equation}
\mathcal{A}(\boldsymbol{\theta},\chi)=
\frac{1}{f_{k}(\chi)}
\frac{\partial \textbf{x}}{\partial \boldsymbol{\theta}}.
\end{equation}
In the limit of weak-field metric (small $\Phi$), the integral in \autoref{sep} can be approximated by considering the series expansion in power of $\Phi$ and truncating the series at the first term.
With this assumptions, given that $\nabla_{\bot}\boldsymbol{\Phi}^0$ not dependent from $\boldsymbol{\theta}$, the Jacobian matrix can be written as:
\begin{equation}
    \mathcal{A}_{ij}(\boldsymbol{\theta},\chi)
     =\delta_{ij}-\frac{2}{c^2}
    \int_0^{\chi} d\chi'
     \frac{f_k(\chi-\chi')f_k(\chi')}{f_k(\chi)}
     \Phi_{ij}(f_k(\chi')\boldsymbol{\theta},\chi').
\end{equation}
This, also known as Born approximation, corresponds to integrating the potential gradient along the unperturbed ray.
If we define the 2D potential, the \textit{lensing potential} as:
\begin{equation}
    \psi(\boldsymbol{\theta},\chi) \equiv
 -\frac{2}{c^2}
= \int_0^{\chi} d\chi'
   \frac{f_k(\chi-\chi')f_k(\chi')}{f_k(\chi)f_k(\chi')}
  \Phi(f_k(\chi')\boldsymbol{\theta},\chi')
\end{equation}
 the Jacobi matrix can be written as:
\begin{equation}
    \mathcal{A}_{ij}=\delta_{ij}-\partial_i \partial_j\psi.
\end{equation}

From the parametrization of the symmetrical matrix $\mathcal{A}$, we can define the spin-two shear $\gamma=(\gamma_1,\gamma_2)$ and the scalar convergence field, $\kappa$. 
Hence, the convergence and the shear are defined as the second derivative of the potential:
\begin{equation}\label{kshear}
    \kappa=\frac{1}{2}(\partial_1\partial_1+\partial_2\partial_2)\psi;  
\end{equation}
\begin{equation}
 \gamma_1=\frac{1}{2}(\partial_1\partial_1-\partial_2\partial_2)\psi; \  \  \
 \gamma_2=\partial_1\partial_2\psi;   
\end{equation}
the two fields $\gamma$ and $\kappa$ describe the distortion in the shape of the image, and the change in the angular size, respectively.
By combining the 2D Poisson equation with the \autoref{kshear}, the convergence $\kappa$ becomes:
\begin{equation}\label{born_approx}
    \kappa_{born}(\boldsymbol{\theta})= \frac{3H_0^2 \Omega_m}{2c^2}
    \int_0^{\chi_s} 
    \frac{d\chi}{a(\chi)}
    g(\chi)
    \delta(f_k(\chi)\boldsymbol{\theta},\chi),
\end{equation}
where we define the \textit{lensing efficiency}:
\begin{equation}
   g(\chi) \equiv
  \int_{\chi}^{\chi_{lim}} \text{d}\chi'
   n(\chi')
    \frac{f_k(\chi'-\chi)}{f_k(\chi')}.
\end{equation}

@@

\subsubsection{Intrinsic alignments (NLA)}

The galaxy ellipticity observed by a telescope can be decomposed in the cosmic shear signal $\gamma$ and the intrinsic ellipticity of the source $\epsilon^{int}$, where the latter is the combination of the alignment term $\epsilon^{IA}$ and the random component $\epsilon^{ran}$.

Different theoretical models have been proposed to describe the physics of Intrinsic Alignments (IA), such as the Non-Linear tidal Alignment model (NLA) (e.g. in \citet{Bridle_2007}), the tidal torquing model \citep{PhysRevD.70.063526, 10.1046/j.1365-8711.2001.04105.x}, or the combination of both the Tidal Alignment and Tidal Torquing model (TATT)  \citep{PhysRevD.100.103506}.
We model the IA effect using the NLA description \citep{harnois2021cosmic}, i.e. assuming a linear coupling between the intrinsic galaxy shapes and the non-linear projected tidal fields $s_{ij}$:
\begin{equation}\label{complex_e}
    \epsilon_{1}^{IA}=- \frac{A_{IA}\bar{C}_1\bar{\rho}(z=0)}{D(z)} (s_{xx}-s_{yy}), \ \ 
      \epsilon_{2}^{IA}=-\frac{2A_{IA}\bar{C}_1\bar{\rho}(z=0)}{D(z)} s_{xy},
\end{equation}
from which the observed ellipticities are computed as:
 \begin{equation}
    \boldsymbol{\epsilon}^{obs}=
    \frac{\boldsymbol{\epsilon}^{int}+\textbf{g}}{1+\boldsymbol{\epsilon}^{int,*}\textbf{g}}, \ \  \text{with}  \ \
    \boldsymbol{\epsilon}^{int}=
    \frac{\boldsymbol{\epsilon}^{
     IA}+\boldsymbol{\epsilon^{ran}}}
     {1+\boldsymbol{\epsilon}^{IA,*}\boldsymbol{\epsilon^{ran}}}.
\end{equation}
The $A_{IA}$ term in \autoref{complex_e} defines the strength of the tidal coupling, $\bar{C}_1$ is a constant
calibrated in \citet{brown2002measurement}, $D(z)$ is the linear growth function and $\bar{\rho}$ is the matter density.

\subsection{Lensing Summary Statistics }\label{stat}
To extract the cosmological information from the simulated $\kappa-$maps, we use two summary statistics for weak lensing observable: the angular power spectrum and the starlet peak counts \citep{lin2016new}.

\subsubsection{Angular Cls
}
Second-order statistics, both in the form of shear 2-point correlation function $\xi_{\pm}(\theta)$, or its counterpart in Fourier space, the angular power spectrum $C_{\ell}$, have been widely used to extract the cosmological information from weak lensing surveys. \
In the Limber approximation, the angular power spectrum of the convergence field for a given tomographic bin can be computed as:
\begin{equation}
    C_{\kappa}(\ell)=\frac{9\Omega_m^2H_0^4}{4c^4}
    \int_0^{\chi_{lim}} \text{d}\chi 
    \frac{g^2(\chi)}{a^2(\chi)}P_{\delta}
    	\left( k=\frac{\ell}{f_K(\chi)},\chi \right),
\end{equation}
where $P_{\delta}$ defines the matter power spectrum of the density contrast.

The Intrinsic Alignment (IA) signal adds an excess correlation to the two-point shear correlation function (also known as cosmic shear GG or shear-shear correlation) with two terms: 1) the intrinsic-intrinsic (II) term, tracing the correlation of the intrinsic shape of two galaxies  and 2) and the intrinsic-shear coupling (GI) term, describing the correlation between the intrinsic ellipticity of one galaxy with the shear of another galaxy \citep{kilbinger2015cosmology}.
% In the Limber approximation, the $C_{\ell}^{II}$ and $C_{\ell}^{GI}$ contributes can be computed from the matter power spectra:
The matter power spectra for the IA terms are defined as:
\begin{align}
    P_{II}(k,z)=& \left ( \frac{A_{IA}\bar{C}_1\bar{\rho}(z)}{\bar{D}(z)} \right)^2 a^{4}(z)P_{\delta}(k,z) 
    \\
    P_{GI}(k,z)=&  \frac{A_{IA}\bar{C}_1\bar{\rho}(z)}{\bar{D}(z)} a^{2}(z)P_{\delta}(k,z),
\end{align}
where $\bar{D}(z) \equiv D(1+z)$ \citep{harnois2021cosmic}.
\denise{Under the Limber
approximation the projected angular power spectra for the IA terms become:
\begin{align}
    C_{II}=& \int_0^{\chi_{lim}} \text{d}\chi 
    \frac{n^2(\chi)}{a^2(\chi)}P_{II}(k,\chi), \\
    C_{GI}= \frac{3\Omega_m H_0^2}{2c^2} & \int_0^{\chi_{lim}} \text{d}\chi 
    \frac{g(\chi)n(\chi)}{a(\chi)}P_{GI}(k,\chi).
\end{align}
}

\subsubsection{Wavelet peak counts}
\paragraph{Wavelet Transform} The wavelet transform has been widely used in analysing astronomical images due to its ability to decompose astronomical data into components at different scales. 
This multiscale approach is well-suited for the study of astronomical data since their complex hierarchical structure. \denise{A wavelet function $\psi(x)$ is a function that satisfies the admissibility condition:
\begin{equation}
    \int_{\mathbb{R}^{+}} |\hat{\psi}(k)|^2 \frac{dk}{k}<\infty,
\end{equation}
where we indicate with $\hat{\psi}(k)$ the Fourier transform of $\psi(x)$, with $\int\psi(x)dx=0$ in order to satisfy the admissibility condition. 
A given signal is decomposed in a family of scaled and translated functions:
\begin{equation}
    \psi_{a,b}(x)=\frac{1}{\sqrt{a}}\psi \left( \frac{x-b}{a}\right),
\end{equation}

where $\psi_{a,b}$ is the so-called \textit{daughter wavelets}, scaled and translated version of the \textit{mother} wavelet, with $a$ and $b$ scaling and translation parameters.
The continuous wavelets transform is defined from the projections of a function $f \in L_2(\mathbb{R})$ onto the family of daughter wavelets. The coefficients of this projection represent the wavelet coefficient, obtained by :
\begin{equation}
    W_f(a,b)=\int_{\mathbb{R}} f(x)\psi^{*}_{a,b}(x)dx=\frac{a}{\sqrt{a}}\int_{\mathbb{R}}f(x)\psi^{*}\left( \frac{x-b}{a}\right)dx ,
\end{equation}
with $\psi^{*}$ the complex conjugate of $\psi$, and $ \forall a \in \mathbb{R}^{+}$, $ b \in \mathbb{R.}$  }
In this work, we filter the original convergence maps with the starlet transform, an isotropic and undecimated (i.e. not down-sampled)  wavelet transform, suited for astronomical applications where objects are mostly more or less isotropic \citep{4060954}.

It decomposes an image $c_0$ as the sum of all the wavelet scales and the coarse resolution image $c_J$:
\begin{equation}\label{wav_des}
    c_0(x,y)=c_J(x,y)+\sum_{j=1}^{J_{max}} w_j(x,y)
\end{equation}
where $J_{max}$ is the maximum number of scales and $w_j$ is the wavelet images showing the details of the original image at dyadic scales with a spatial size of $2^j$ pixels and $j = J_{max} + 1$.

The starlet wavelet function is a specific translational invariant wavelet transform:
\begin{equation}
   \frac{1}{4} \Psi \left ( \frac{x_1}{2}, \frac{x_2}{2} \right )=
    \phi (x_1,x_2)- \frac{1}{4} \phi  \left ( \frac{x_1}{2},  \frac{x_2}{2} \right )
\end{equation}
specified by an isotropic scaling function $\phi$, that,  for astronomical application, can be defined as a B-spline of order 3:
\begin{equation}
\phi_{1D}(x)= \frac{1}{12} (|x-2|^3-4|x-1|^3+6|x|^3-4|x+1|^3+|x+2|^3 ).
\end{equation}
\denise{
The N-dimensional scaling functions can be built starting from the separable product of N $\phi_{1D}$: $\phi(x_1,x_2)=\phi_{1D}(x_1)\phi_{1D}(x_2)$. }
Each set of wavelet coefficients $w_j$ is obtained as the convolution of the input map with the corresponding wavelet kernel. For a full description of the starlet transform function, see \citet{4060954} and \citet{10.5555/1830428}.
